---
title: "Lab 1: R Basics"
subtitle: "DS 01-101: In-Depth Introduction to Data Science (Spring 2026)"
author: "Hung Tong"
institute: "Rowan University"
format: 
  html:
    self-contained: true
    preview-links: true
editor: visual
toc: true
echo: true
eval: false
output: true
toc-expand: 2
---

## Reading

From **R Coding Basics: An Introduction to the Basics of Coding in R** by Dr. Gaston Sanchez:

-   First Contact with R

-   Getting Help

## Topics

-   Mathematical operations

-   Operation precedence

-   Comments

-   Objects, variables, and assignments

-   Variable names and reserved words

-   Functions

-   Help Documentation

-   Packages

## R as a Calculator

Write R code to evaluate each expression:

1.  $\dfrac{5}{8} + \dfrac{1}{4} - \dfrac{5}{6}$

2.  $\dfrac{2 - 79}{-11}$

3.  $\dfrac{2}{5} \times \dfrac{7}{8}$

4.  $\left( -\dfrac{2}{3} \right) \times \dfrac{5}{8} \times \dfrac{3}{4}$

5.  $3\left( -\dfrac{5}{12} + \dfrac{3}{8} \right)$

6.  $-6^2 - [7 + (-2)^3]$

7.  $3^{-2} + 4^{-1}$

8.  $\dfrac{7^5 + 1}{7^5 - 1}$

9.  $\dfrac{-4 + \sqrt{12}}{4}$

10. $-\left \lvert \dfrac{1}{3} \right \rvert - \left \lvert -\dfrac{2}{3} \right \rvert$

## x, y, z

Evaluate the expression at $x = 3$, $y = -2$, and $z = 4$

1.  $x + 3y + z$

2.  $6z + 5x - 3y$

3.  $x^2 - 5y + 4z$

4.  $z^2 + 6y - x$

5.  $\dfrac{x - y}{5z}$

6.  $\dfrac{4z - 2y}{20 x}$

## Balance in an Account

The balance $A$ in an account that earns an annual compound interest rate $r$ (in decimal form) for $t$ years is given by

$$
A = P \left( 1 + \dfrac{r}{n} \right)^{nt},
$$

where $P$ is the principal (initial deposit) and $n$ is the number of compoundings per year.

1.  We deposit $\$6000$ in an account with an annual interest rate of $1.5\%$. What is our balance after $5$ years when the interest is compounded quarterly ($n = 4$)?

2.  We deposit $\$10,000$ in an account with an annual interest rate of $3.75\%$. What is our balance after $12$ years when the interest is compounded daily ($n = 365$)?

## The First 100k is the Hardest

The compound interest formula can be rewritten as

$$
P = \dfrac{A}{\left( 1 + \dfrac{r}{n} \right)^{nt}}
$$

Find the principal $P$ that we need to invest today at $4.5\%$ annual interest compounded monthly to earn $\$100,000$ in the next $20$ years.

## Comic Book

The demand function for a limited edition comic book is given by

$$
p = 3000 \left( 1 - \dfrac{5}{5 + e^{-0.015x}} \right)
$$

Find the price $p$ for a demand of $x = 75$ units and $x = 200$ units.

## Area of a Circular Cylinder

The formula for the volume of a cylinder is

$$
V = \pi r^2 h
$$

where $r$ is the radius and $h$ is the height.

Find the height of a cylindrical shipping container with a volume of $15,607$ cubic inches and a radius of $12$ inches. You may find reading `help(pi)` or `?pi` useful.

## R as a Graphing Calculator

The `plot()` function provides various graphical displays in R. For example, to plot the following linear function

$$
f(x) = 5 - 3x
$$

we can use the following code

```{r}
#| eval: true

fx <- function(x) {
  5 - 3 * x    # modify the function
}

plot(fx, from = -5, to = 10)    # adjust from = ... and to = ...
```

To plot the following quadratic function

$$
f (x) = x^2 + 2x - 3
$$

we can use the following code

```{r}
#| eval: true

fx <- function(x) {
  x^2 + 2 * x - 3    # modify the function
}

plot(fx, from = -4, to = 4)    # adjust from = … and to = …
```

Modify the above R code to plot the following functions:

1.  $f (x) = x^3 - 3 x^2$

2.  $f (x) = x^{2/3}$

3.  $f (x) = x \sqrt{x + 3}$

4.  $f (x) = \lvert x + 1 \rvert + \lvert x - 1 \rvert$

## Monthly Payment

The length $t$ in years of a home mortgage of $\$150,000$ at $8\%$ interest can be approximated by the model

$$ t = 12.542 \ln \left( \dfrac{x}{x - 1} \right), \quad \text{for $x > 1$} $$

where $x$ is the monthly payment in thousand dollars.

1.  Approximate the length of this mortgage when the monthly payment is $\$1254$.

2.  Approximate the total amount paid over the term of this mortgage with a monthly payment of $\$1254$.

3.  Approximate the total interest charge for a monthly payment of $\$1254$.

4.  Plot the model and describe the relationship between monthly payment and length of the given mortgage.
