---
title: "Lecture 1: R Basics"
author: "Hung Tong"
institute: "Rowan University"
format: 
  revealjs:
    slide-number: true
    self-contained: true
    code-overflow: wrap
editor: visual
echo: true
eval: false
output: true
code-line-numbers: false
---

## Reading

From **R Coding Basics: An Introduction to the Basics of Coding in R** by Dr. Gaston Sanchez:

-   First Contact with R

-   Getting Help

## Topics

-   Mathematical operations

-   Operation precedence

-   Comments

-   Objects, variables, and assignments

-   Variable names and reserved words

-   Functions

-   Help Documentation

-   Packages

## Mathematical Operations

-   Common mathematical operations in R include:

```{r}

2 + 2     # Add

3 - 4     # Subtract

5 * 2     # Multiply

4 / 2     # Divide

2 ^ 3     # Exponent
2 ** 3    # Exponent as well
```

## Operation Precedence

-   In R, operator precedence (from highest to lowest) is:

```{r}

# Parenthesis
(2 + 3) * 4

# Exponent
2 ^ 3
2 ** 3

# Multiply and divide (from left to right)
5 * 2
4 / 2

# Add and subtract
2 + 2
3 - 4
```

## Your Turn

Evaluate the following expressions in R:

-   $19 - 3 \times 6$

-   $20 - 2 \times 3^2 + 11$

-   $3 + 8 \div 2 \times 2$

-   $7 - [5 \times 3 + 2^3]$

## Answer

```{r}

19 - 3 * 6

20 - 2 * 3^2 + 11

3 + 8 / 2 * 2

7 - (5 * 3 + 2^3)
```

## Your Turn

Add parentheses `()` to each expression so that the output matches the comment.

```{r}

5 * 2 + 4 / 2    # 15

5 * 2 + 4 / 2    # 7

5 * 2 + 4 / 2    # 20
```

## Answer

```{r}

5 * (2 + 4) / 2    # 15

(5 * 2 + 4) / 2    # 7

5 * (2 + 4 / 2)    # 20
```

## Comments

-   Comments are notes to explain what the code does.

-   In R, comments are created using a hash `#`.

-   On a line, everything after a hash will be ignored. In other words, even if the comments contain some code, it will not be executed.

-   The keyboard short cut to comment out blocks of code is `Command` or `Ctrl` + `Shift` + `c`

## Object, Variable, and Assignment

-   An object is anything that stores values or instructions.

-   A variable is a name associated to an object so that we can access the object in memory.

-   A variable can be created using the assignment operator `<-` (left arrow) or the equal sign `=`.

```{r}

# 100 is an object itself
100

# We name this object 'eagles'
eagles <- 100

# The variable 'eagles' allows us to access the object 100
eagles
```

## Variable Names

-   Variable names are case-sensitive.

```{r}

# These names are NOT the same
eagles

Eagles

EAGLES

eaGleS

EagLEs

EaGlEs
```

## Variable Names

-   A variable name must start with a letter or a period (not recommended). The rest can only consist of letters `a-z, A-Z`, numbers `0-9`, underscores `_`, and periods `.`

-   A variable name cannot be a reserved word (`TRUE`, `FALSE`, `if`, `for`, `while`, etc.)

```{r}

# Valid variable names
age            
height_cm1    
weight2.kg    

# Valid but not recommended
.data       
True
```

## Reserved Words

-   Reserved words are words with special meaning and cannot be used as variable names.

```{r}

if, else, for, while, function 

break, stop, continue, next

TRUE, FALSE

NULL, NA, Inf, NaN
```

## Your Turn

The following R code attempts to create variables. Run each line and explain why the variable name is invalid.

```{r}

9null <- 5

null%9 <- 5

null 9 <- 5

_null9 <- 5 

null,9 <- 5

NULL <- 5
```

## Answer

```{r}

9null <- 5     # A variable name must start with a letter
_null9 <- 5    

null%9 <- 5    # The rest can only be letters, numbers, +, and periods
null 9 <- 5
null,9 <- 5    

NULL            # A variable name cannot be a reserved word
```

## Naming Styles

Three common naming styles for variables include

```{r}

eagles_green_swag    # snake case

eaglesGreenSwag      # camel case

eagles.green.swag    # period case
```

## **Functions**

-   A function is a collection of R code to perform a complex task or obtain certain results.

-   R has many useful built-in functions.

```{r}

abs(-7)         # absolute value

sqrt(25)        # squared root

exp(3)          # exponential

log(8)          # natural log

factorial(6)    # 1 x 2 x 3 x 4 x 5 x 6
```

## Functions

```{r}

# Absolute value, sign function, and squared root
abs, sign, sqrt

# Rounding of numbers
round, floor, ceiling, trunc, signif

# Logarithms and Exponentials
exp, log, expm1, log1p

# Combination and factorial
choose, factorial, lchoose, lfactorial

# Trigonometric and hyperbolic functions
cos, sin, tan, cospi, sinpi, tanpi, acos, asin, atan
cosh, sinh, tanh, acosh, asinh, atanh

# Beta and gamma functions
beta, lbeta, gamma, lgamma, digamma, trigamma, psigamma
```

## Getting Help

-   If we want to learn more about a function, we can refer to its help documentation.

```{r}

# Using the operator '?'
?abs
?sqrt

# Using the function help()
help(abs)
help(sqrt)
```

## Your Turn

Use the help documentation to explain the purpose of the following functions.

```{r}

sign(-17)

log1p(3)

choose(5, 2)
```

## Answer

```{r}

sign(-17)       # return 1 if x is positive, -1 if x is negative
#> [1] -1

log1p(3)        # compute log(1 + x) = log(1 + 3)
#> [1] 1.386294

choose(5, 2)    # return binomial coefficient 5 choose 2
#> [1] 10
```

## Your Turn

What does the following code do? Use the help documentation of `round()` to confirm your guess/observation.

```{r}

round(6.267096)

round(6.267096, digits = 1)
round(6.267096, 1)

round(6.267096, digits = 2)
round(6.267096, 2)
```

## Answer

```{r}

# Round to the whole number (digits = 0)
round(6.267096)
#> [1] 6

# Round to 1 decimal place
round(6.267096, digits = 1)
#> [1] 6.3

# Round to 2 decimal places
round(6.267096, digits = 2)
#> [1] 6.27
```

## Packages

-   A package is a collection of functions and data for specific purpose.

-   Most developers make their packages freely available on the Comprehensive R Archive Network (CRAN).

```{r}

# Install a package
install.packages("bannerCommenter")

# Load an installed package
library(bannerCommenter)
library("bannerCommenter")

# Uninstall a package
remove.packages("bannerCommenter")
```
